---
title: "Schedule"
sidebar: false
toc: false
page-layout: full
---

```{r packages-data, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

withr::with_dir(here::here(), {
  schedule_nested <- targets::tar_read(schedule_page_data)
  targets::tar_load(schedule_ical_file)
})

base_url <- yaml::read_yaml(here::here("_quarto.yml"))$website$`site-url`
ical_url_absolute <- paste0(base_url, "/", schedule_ical_file)
```

::: {.schedule-page}

Here's your roadmap for the summer camp!

- [**Lessons**](/content/index.qmd) ({{< fa book-open-reader >}}): This page contains the slides for the topic. Read and go through these first.

- [**Hands-on Sessions**](/handson/index.qmd) ({{< fa laptop-code >}}): This page provides the interactive, practical exercises that correspond with each lesson. Engage with these sessions after reviewing the slides to solidify your understanding. We will go through these during each odd numbered session.

- [**Labs**](/labs/index.qmd) ({{< fa chalkboard-teacher >}}): This page contains, when needed, interactive sessions, and also the contents for each lab. Go through these after doing the lessons and hands-on sessions. This is more on your own, so we will go through these during each even numbered session.

:::::: {.row .d-flex .justify-content-center}
::::: {.col-12 .col-sm-10 .col-md-6}
:::: {.card .bg-danger .text-white}
::: {.card-body}

**tl;dr**: You should follow this general process for each session:

- Read the slides in the lesson page ({{< fa book-open-reader >}})
- Work through the hands-on page ({{< fa laptop-code >}})
- Do the labs page ({{< fa chalkboard-teacher >}})

:::
::::
:::::
::::::

---

::: {.callout-tip}

## Subscribe!

You can subscribe to this calendar URL in Outlook, Google Calendar, or Apple Calendar:

::: {.grid}

::: {.g-col-12 .g-col-md-2}
<p class="download-btn"><a class="btn btn-danger btn-sm" href="/`r schedule_ical_file`" role="button">{{< fa calendar-alt >}} Download</a></p>
:::

::: {.g-col-12 .g-col-md-10}
<p><input class="form-control" type="text" value="`r ical_url_absolute`" onclick="this.select()" readonly></p>
:::

:::

:::

```{r build-table, include=FALSE}
show_table <- function(group_id) {
  # Add a heading
  cat(as.character(paste("\n\n###", schedule_nested$group[[group_id]], "\n\n")))

  # Make the table
  tbl <- schedule_nested$data[[group_id]] |>
    kbl(escape = FALSE, align = "rlccc", table.attr = 'class="schedule-table"') |>
    kable_styling() |>
    column_spec(1, width = "20%", extra_css = "padding-right: 20px;") |>
    column_spec(2, width = "48%") |>
    column_spec(3:5, width = "8%")

  cat(tbl)
}
```

```{r show-table, echo=FALSE, results="asis"}
walk(seq(1, nrow(schedule_nested)), ~show_table(.x))
```

:::
