---
title: "Schedule: Summer 2024"
sidebar: false
toc: false
page-layout: full
---

```{r packages-data, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

withr::with_dir(here::here(), {
  schedule_nested <- targets::tar_read(schedule_page_data)
  targets::tar_load(schedule_ical_file)
})

base_url <- yaml::read_yaml(here::here("_quarto.yml"))$website$`site-url`
ical_url_absolute <- paste0(base_url, "/", schedule_ical_file)
```

::: {.schedule-page}

Here's your roadmap for the math camp!

The content for each week is organized into three sections, which you'll follow in this order:

- [**Lessons**](/2024/weeks/week01/page.qmd) ({{< fa book-open-reader >}}): Begin with the Lesson section. These pages contain the slides for each topic. Make sure to read through them first to build your foundation for the week.

- [**Hands-on Sessions**](/2024/weeks/week01/hands-on.qmd) ({{< fa chalkboard-teacher >}}): Next, move on to the Hands-on Sessions. These pages include practical exercises that will help you apply what you've learned in the lessons. We'll work on these together during the odd-numbered sessions each week.

- [**Labs**](/2024/weeks/week01/lab.qmd) ({{< fa laptop-code >}}): Finally, explore the Lab section. These pages offer additional interactive components and lab exercises. You'll complete these primarily on your own, and we'll discuss them during the even-numbered sessions each week.

Each week's content is structured in this same way, so be sure to follow this sequence to get the most out of your learning experience.

:::::: {.row .d-flex .justify-content-center}
::::: {.col-12 .col-sm-10 .col-md-6}
:::: {.card .bg-danger .text-white}
::: {.card-body}

**tl;dr**: You should follow this general process for each session:

- Read the slides in the lesson page ({{< fa book-open-reader >}})
- Work through the hands-on page ({{< fa chalkboard-teacher >}})
- Do the labs page ({{< fa laptop-code >}})

:::
::::
:::::
::::::

---

::: {.callout-tip}

## Subscribe!

You can subscribe to this calendar URL in Outlook, Google Calendar, or Apple Calendar:

::: {.grid}

::: {.g-col-12 .g-col-md-2}
<p class="download-btn"><a class="btn btn-danger btn-sm" href="/`r schedule_ical_file`" role="button">{{< fa calendar-alt >}} Download</a></p>
:::

::: {.g-col-12 .g-col-md-10}
<p><input class="form-control" type="text" value="`r ical_url_absolute`" onclick="this.select()" readonly></p>
:::

:::

:::

```{r build-table, include=FALSE}
show_table <- function(group_id) {
  # Add a heading
  cat(as.character(paste("\n\n###", schedule_nested$group[[group_id]], "\n\n")))
  
  # Make the table
  tbl <- schedule_nested$data[[group_id]] %>% 
    select(-subgroup) %>% 
    kbl(escape = FALSE, align = "rlccc", table.attr = 'class="schedule-table"') %>% 
    kable_styling() %>% 
    column_spec(1, width = "15%", extra_css = "padding-right: 20px;") %>% 
    column_spec(2, width = "50%") %>% 
    column_spec(3:5, width = "10%") %>% 
    pack_rows(
      index = schedule_nested$subgroup_index[[group_id]],
      label_row_css = "border-bottom: 1px solid #adb5bd;"
    )
  
  cat(tbl)
}
```

```{r show-table, echo=FALSE, results="asis"}
walk(seq(1, nrow(schedule_nested)), ~show_table(.x))
```

:::
