---
title: "Streamlining Research with R"
author: "Rony Rodrigo Maximiliano Rodriguez-Ramirez"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "libs"
    chakra: "libs/remark-latest.min.js"
    css: ["default", "css/ath-slides.css", "css/ath-inferno-fonts.css", "css/animate.css"]
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
      # beforeInit: "libs/offline-mathjax.js"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.retina = 3, fig.align = "center")
```

```{r packages-data, include=FALSE}
library(tidyverse)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "share_again"))
```

class: center middle main-title section-title-4

# Streamlining Research with R

.class-info[

**Lesson 3**

.light[API 209: Advanced Quantative Methods<br>
TF: Rony Rodriguez-Ramirez<br>
Summer 2024]

]

---

name: outline
class: title title-inv-5

# Plan for today

1. List
2. List
3. List
4. List

---

layout: false
name: knowing-each-other
class: center middle section-title section-title-crimson

# First Part

---

layout: true
class: title title-crimson

---

# First Slide


---

# Second Slide


---

layout: false
name: knowing-each-other
class: center middle section-title section-title-crimson

# Second Part

---

layout: true
class: title title-crimson

---

# First Slide


---

# Second Slide

