---
title: "Mastering Data Manipulation in R"
date: "2024-08-14"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

## Overview

This session will focus on advanced data manipulation techniques using R and the tidyverse. While the primary resource for this session will be the slides, consider reflecting on the following questions as you engage with the material.

## Guiding Questions

::: {.callout-warning}

#### These are optional!

You don't need to answer all of these—or even any of them! They are here to help guide your thinking and reflection on the content. Feel free to explore whatever aspects you find most interesting.
:::

- How do complex filtering and selection operations enhance data analysis?
- What are the benefits of using the tidyverse for data manipulation?
- How do different data reshaping techniques (e.g., pivoting) impact the way data is analyzed and interpreted?
- What challenges do you encounter when working with large datasets, and how can they be overcome with efficient R techniques?
- How does the native pipe (`|>`) improve code readability and workflow in R?

## Readings

- No readings for this session.

## Slides

The slides for this session are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also navigate through the embedded slides directly on this page using your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/01-slides")

xaringanExtra::embed_xaringan(
  "/slides/01-slides.html",
  ratio = "16:9",
  border = "none",
  margin = "0 auto"
)
```

:::{.callout-tip}
**Pro tip**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while viewing the slides, you can access a list of special commands to enhance your experience.
:::
