---
title: "Streamlining Research with R"
date: "2024-08-26"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

## Overview

IN CONSTRUCTION!


## Guiding Questions

::: {.callout-warning}

#### These are optional!

You don't need to answer all of these—or even any of them! They are here to help guide your thinking and reflection on the content. Feel free to explore whatever aspects you find most interesting.
:::

- Question 1

## Readings

- No readings for this session.

## Slides

The slides for this session are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also navigate through the embedded slides directly on this page using your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/04-slides")

xaringanExtra::embed_xaringan(
  "/slides/04-slides.html",
  ratio = "16:9",
  border = "none",
  margin = "0 auto"
)
```

:::{.callout-tip}
**Pro tip**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while viewing the slides, you can access a list of special commands to enhance your experience.
:::
