---
title: "🗓️ Week 02 - Tidy Data and Visualization"
date: "2026-08-20"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(if (file.exists("R/slide-things.R")) "R/slide-things.R" else "../../../R/slide-things.R")
```

## Overview

In this week, we dive into the concepts of tidy data and data visualization. You'll learn how to organize your data in a format that is easy to manipulate and visualize. We will focus on transforming data into a "tidy" format where each variable is a column, each observation is a row, and each type of observational unit forms a table. Additionally, we'll cover essential visualization techniques using `ggplot2` to explore and communicate your findings effectively.

## 👨‍🏫 Lesson Slides

Either click on the slide area below or click [here](slides02.qmd) to view it in fullscreen.
Use your keypad to navigate the slides.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("slides02")
```

<div class="container" style="position: relative;overflow: hidden;width: 100%;min-height:600px;">
<iframe src="slides02.html" style="position=absolute;top: 0;left: 0;bottom: 0;right: 0;width: 100%;min-height:600px;"></iframe>
</div>

## Guiding Questions

::: {.callout-warning}
#### These are optional!
You don't need to answer all of these—or even any of them! They are here to help guide your thinking and reflection on the content. Feel free to explore whatever aspects you find most interesting.
:::

- What are the key principles of tidy data, and why are they important?
- How does the structure of your data impact the type of visualizations you can create?
- In what ways can data visualization help in uncovering trends and patterns?

## Readings

- No readings for this session.

## Lesson Focus

This week is less about memorizing syntax and more about building analytical judgment:

- how to decide whether data are tidy enough for analysis;
- how plot choices shape interpretation;
- how to communicate a pattern clearly without overstating it;
- how to move from a rough exploratory graphic to a plot that belongs in a memo or problem set.
