---
title: "🗓️ Week 01 - Introduction, Context & Data Wrangling"
date: "2024-08-15"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

::: {.callout-warning}

#### STOP

> If you didn't complete the **summer assignments**, you should definitely make time to do complete the following primers. The original content is coming from RStudio and was adapted by Prof. Andrew Heiss.

For the first part of this week's lesson, you need to work through a few of Posit's introductory primers. You'll do these in your browser, where you can type code and see results immediately.

You'll learn some of the basics of R, as well as some powerful methods for manipulating data with the {dplyr} package.

Complete these primers. It may seem like there are a lot, but they're short and go fairly quickly, especially as you get the hang of the syntax. Also, **I have no way of seeing what you do or what you get wrong or right**, and that's totally fine! If you get stuck or want to skip some (or if it gets too easy), feel free to move on!

- **The Basics**
    - [Visualization basics](https://r-primers.andrewheiss.com/basics/01-visualization-basics/)
    - [Programming basics](https://r-primers.andrewheiss.com/basics/02-programming-basics/)
- **Work with Data**
    - [Working with tibbles](https://r-primers.andrewheiss.com/transform-data/01-tibbles/)
    - [Isolating data with {dplyr}](https://r-primers.andrewheiss.com/transform-data/02-isolating/)
    - [Deriving information with {dplyr}](https://r-primers.andrewheiss.com/transform-data/03-deriving/)

The content from these primers comes from the (free and online!) book [*R for Data Science* by Garrett Grolemund and Hadley Wickham](https://r4ds.hadley.nz/). I highly recommend the book as a reference and for continuing to learn and use R in the future (like running regression models and other types of statistical analysis).

:::

## Introduction

In this first week, we will cover what you can expect to learn from this course and the course logistics: all you need to know about the structure of the lectures, classes, assessments, and how we will interact throughout this course. 

## 👨‍🏫 Lesson Slides 

Either click on the slide area below or click [here](slides.qmd) to view it in fullscreen.
Use your keypad to navigate the slides.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("slides01")
```

<div class="container" style="position: relative;overflow: hidden;width: 100%;min-height:600px;">
<iframe src="slides01.html" style="position=absolute;top: 0;left: 0;bottom: 0;right: 0;width: 100%;min-height:600px;"></iframe>
</div>

<!-- {{< revealjs "slides.html" class="ratio ratio-16x9">}} -->